|

### 2021-11-04 article/article.Rmd
adding para theory and para intro

Regardless of eventual differences in electoral rules, social composition, and internal procedures, all legislatures share one feature: they are collective bodies. As such, the number of representatives in parliament has important consequences for whole polity. Indeed, legislature size is known to have a significant effect on collective action strategies, the bargaining power between state branches, government spending, and representation \citep{crain1979cost, freire2021legislature, PrimoSnyder2008, rogers2002free, Weingastetal1981}. However, how legislature size impacts social welfare remain less understood. On the one hand, enlarged legislatures may increase government spending and lawmaking, which can lead to more public goods \citep{Weingastetal1981}. On the other hand, adding a legislator may increase the number of veto players and exacerbate collective action problems, thus hindering service provision \citep{gottlieb2019countervailing, tsebelis2002veto}. Therefore, what is the effect of larger legislatures on public service provision?

@@

This paper incorporates partisanship and mayoral first-move advantage to a \citet{baron1989bargaining} bargaining model. Mayors make the initial policy proposal that consists of an investment in public goods provision, a decision to extract rents for personal gains, and a vector of side-transfers for councilors. If they fail to secure majoritarian support from councilors, a reversal stage occurs, where the decision leaves the mayor's office and councilors propose the public investments. In this mayor-council bargaining process, costs come from the side-transfers required to secure a majority among councilors. We show that in a model without partisan concerns, bargaining costs always increase with larger city councils, which in turn lower public goods provision and rents. However, partisanship decreases bargaining costs when the chances of electing councilors aligned with the mayor are sufficiently high. Lower bargaining costs release funds for mayors to invest in public goods provision. In this case, a larger city council may have more partisan councilors aligned with the mayor, reducing the mayor's bargaining costs and increasing the equilibrium public goods provision.

We test these theoretical predictions by exploiting an exogenous variation in Brazilian city council sizes. In 2004, the judiciary reinterpreted the Brazilian Constitution and unexpectedly changed the number of councilors in several municipalities. Until then, towns could freely determine their legislature size, but in March 2004 the Supreme Electoral Court created population thresholds and added one councilor to each set of 47,619 inhabitants. Around these cutoffs, the decision represented an exogenous change in city council size, allowing us to investigate the effects of increasing the number of local legislators in Brazil.

@@

Higher mayor support and more appointed bureaucrats correlate with city councilors' service provision expectations and legislation approved in Brazilian municipalities. We surveyed 174 former councilors who served during the 2005-2008 term, and they corroborate that mayors use bureaucratic appointments and political favors to secure support. As the formal model shows, bargaining with councilors requires offering transfers to the representatives. By investigating what councilors prefer to provide to voters, we find that councilors mainly demand job appointments, as these increase their reelection chances and improve municipal welfare. These findings align with a novel dataset we built with 346,553 laws passed in 63 municipalities within ten thousand inhabitants away from the cutoffs. The legislation in municipalities on the left and on the right-hand side of the cutoffs mentioned local public goods. However, municipalities with larger city councils had around 15 percent more public goods provision proposals, what is in line with our expectations.

Finally, we show that the increase in legislature size significantly impacts municipal education and health care. Increasing the city council size by one legislator lowers infant mortality by 2.01 children per one thousand infants born and reduces postneonatal mortality by 0.90 children per one thousand infants that survived their first 28 days. A larger council size also increased enrollment by 2.58 children in elementary education classrooms without significantly diminishing elementary school quality. These services are crucial for welfare and result from an eased bargaining between the mayor and the city council.

This paper contributes to three strands of the literature. First, this paper dialogues with a large literature on multilateral bargaining. Repeated models of multilateral bargaining started with \citet{baron1989bargaining}, who extended the bilateral bargaining game proposed by \citet{rubinstein1982perfect} to more than two players. Several papers followed \citet{baron1989bargaining}, manipulating endogenous recognition power, the size of the required majority, and partisanship in the models \citep{agranov2020persistence, baranski2020determinants, calvert2005legislative, choate2019partisan}.\footnote{See \citet{eraslan2019legislative} for a comprehensive review of multilateral bargaining and \citet{ray2015coalition} for a review on coalition formation.} In this paper, we extend the \citet{baron1989bargaining} model under closed rules and add the mayor's first move and political partisanship. Partisanship here is different from \citet{choate2019partisan} since we consider partisanship as a zero-sum effect that is positive for politicians aligned with the mayor and negative otherwise. This paper investigates the often overlooked effect of legislature size on the bargaining equilibrium strategies.

Second, this paper contributes to the literature on the political and economic effects of larger legislatures. Following the contribution by \citet{Weingastetal1981}, the "law of $1/n$" posits that larger legislatures increase the legislative pressure for providing targeted public goods to local supporters at the expense of the entire polity \citep{freire2021legislature}. The corollary of the "law of $1/n$" is that increasing legislature size promotes excessive public spending.\footnote{\citet{PrimoSnyder2008} show that legislature size may result in positive or negative expenditures, depending on factors such as the degree that local public goods can exclude non-locals or deadweight costs associated with taxation.} In the US metaphor, this would explain significant infrastructure spending in smaller districts. However, the "law of $1/n$" literature rarely mentions the relationship between public goods provision and welfare. This paper is one of the first to study how the number of local legislators affects welfare. Our model provides a theoretical expectation based on the chances of electing a politician aligned with the mayor. The empirical estimates confirm our theoretical predictions, with the additional fact that legislature size has a null effect on expenditures in Brazil.

Finally, our article also contributes to a growing literature on municipal partisan politics. Most papers in this area estimate the effects of the mayor's party on political outcomes. For instance, \citet{gerber2011mayors} and \citet{de2016mayoral} find differences in governance when comparing Democratic and Republican mayors. Specifically about Brazil, \citet{gouvea2021partisanship} shows that social policy expenditure increase with left-wing mayors. \citet{frey2020larger}, using discontinuities in city council size after the 2009 constitutional amendment that reverted the thresholds we study here, shows that changes in council size lower the support for candidates backed by the municipal mayors. This paper reinforces the view that partisanship matters locally, even in countries with weak ideological linkage.

@@

Several cities in the world are governed by the mayor-council system. In mayor-council cities, the power is divided between a mayor that controls the Executive branch and a city council that supports or sanctions the policies proposed by the mayor. In this setting, consider a strategic interaction between a mayor ($M$) and $N \geq 3$ city councilors indexed in $i \in \{1, 2, \cdots, N \}$. The mayor has the prerogative of proposing a vector of policies on which city councilors vote. The provision of public goods is then a bargain between the mayor and the city council. Policy proposals are a combination of public goods provision $g$, rents for the mayor $r$, and a vector of transfer for city councilors \textbf{x}.

@@

We now consider a situation that combines both partisan and non-partisan legislature concerns. Each councilor has a party affiliation. Party affiliations are mutually exclusive, and if a city councilor is aligned with the mayor's party, then she belongs to the government coalition $G \subset \{1, 2, \cdots, N \}$. Otherwise, the city councilor belongs to the opposition $O = \{1, 2, \cdots, N \} \setminus G$. The mayor's policy generates a political value of $p > 0$ for the councilors, and whether it adds or subtracts from the councilor's utility depends on the party affiliation of the councilor. If councilors reject the mayor's proposal, no public goods are provided ($p=0$). Moreover, in equilibrium, we assume that city councilors would prefer to receive transfers instead of providing public goods.\footnote{In the Supplementary Materials, we consider a version of the model where public goods provision depends on the party affiliation of the proposer so that it affects choices in any potential history of the game. The equilibrium public goods provision level is even higher than in the model presented here, as city councilors prefer to approve the mayor's proposal than run the risk of an opposition member becoming the proposer.}

@@

Second, more seats could increase the competitiveness in the election, making politicians provide more services \citep{gottlieb2019countervailing}. In a political system with around thirty parties and a sample average of almost ten parties in the municipalities we study, candidates per seat can increase exponentially. According to the Brazilian electoral law, each party can sponsor 1.5 times the number of seats as candidates. If a party forms a pre-electoral coalition, this number increases to two times the number of seats \citep{coneglian2008lei}. An election in a small town, with ten councilors, may end up with 600 candidates: (30 parties) x (two times the number of seats) x (ten seats). Therefore, an alternative hypothesis to explain the increased service provision is that larger council size increase the competitiveness of the election.\footnote{Despite the common assumption that competitiveness improve welfare, \citet{boulding2014political} and \citet{gottlieb2019countervailing} can increase budgetary pressure, lower social policy, lead to bargaining inefficiencies.}

@@

\pagebreak

>>>>

Regardless of eventual differences in electoral rules, social composition, and internal procedures, all legislatures share one feature: they are collective bodies. As such, the number of representatives has important consequences for the whole polity. Indeed, legislature size is known to have a significant effect on collective action strategies, the bargaining power between state branches, government spending, and representation \citep{crain1979cost, freire2021legislature, PrimoSnyder2008, rogers2002free, Weingastetal1981}. However, how legislature size impacts social welfare remains less understood. On the one hand, enlarged legislatures may increase government spending and lawmaking, which can lead to more public goods \citep{Weingastetal1981}. On the other hand, adding a legislator may increase the number of veto players and exacerbate collective action problems, thus hindering service provision \citep{gottlieb2019countervailing, tsebelis2002veto}. Therefore, what is the effect of larger legislatures on public service provision?

@@

This paper incorporates partisanship and mayoral first-move advantage to a \citet{baron1989bargaining} bargaining model. Mayors make the initial policy proposal that consists of an investment in public goods provision, a decision to extract rents for personal gains, and a vector of side-transfers for councilors. If they fail to secure majoritarian support from councilors, a reversal stage occurs, where the decision leaves the mayor's office, and councilors propose the public investments. In this mayor-council bargaining process, costs come from the side-transfers required to secure a majority among councilors. We show that in a model without partisan concerns, bargaining costs always increase with larger city councils, which in turn lower public goods provision and rents. However, partisanship decreases bargaining costs when the chances of electing councilors aligned with the mayor are sufficiently high. Lower bargaining costs release funds for mayors to invest in public goods provision. In this case, a larger city council may have more partisan councilors aligned with the mayor, reducing the mayor's bargaining costs and increasing the equilibrium public goods provision.

We test these theoretical predictions by exploiting an exogenous variation in Brazilian city council sizes. In 2004, the judiciary reinterpreted the Brazilian Constitution and unexpectedly changed the number of councilors in several municipalities. Until then, towns could freely determine their legislature size, but in March 2004, the Supreme Electoral Court created population thresholds and added one councilor to each set of 47,619 inhabitants. Around these cutoffs, the decision represented an exogenous change in city council size, allowing us to investigate the effects of increasing the number of local legislators in Brazil.

@@

Higher mayor support and more appointed bureaucrats correlate with city councilors' service provision expectations and legislation approved in Brazilian municipalities. We surveyed 174 former councilors who served during the 2005-2008 term, and they corroborate that mayors use bureaucratic appointments and political favors to secure support. As the formal model shows, bargaining with councilors requires offering transfers to the representatives. By investigating what councilors prefer to provide to voters, we find that councilors mainly demand job appointments, as these increase their reelection chances and improve municipal welfare. These findings align with a novel dataset we built with 346,553 laws passed in 63 municipalities within ten thousand inhabitants away from the cutoffs. The legislation in municipalities on the left and on the right-hand side of the cutoffs mentioned local public goods. However, municipalities with larger city councils had around 15 percent more public goods provision proposals, which is in line with our expectations.

Finally, we show that the increase in legislature size significantly impacts municipal education and health care. Increasing the city council size by one legislator lowers infant mortality by 2.01 children per one thousand infants born and reduces postneonatal mortality by 0.90 children per one thousand infants that survived their first 28 days. A larger council size also increased enrollment by 2.58 children in elementary education classrooms without significantly diminishing elementary school quality. These services are crucial for welfare and result from eased bargaining between the mayor and the city council.

This paper contributes to four strands of the literature. First, this paper dialogues with a large literature on multilateral bargaining. Repeated models of multilateral bargaining started with \citet{baron1989bargaining}, who extended the bilateral bargaining game proposed by \citet{rubinstein1982perfect} to more than two players. Several papers followed \citet{baron1989bargaining}, manipulating endogenous recognition power, the size of the required majority, and partisanship in the models \citep{agranov2020persistence, baranski2020determinants, calvert2005legislative, choate2019partisan}.\footnote{See \citet{eraslan2019legislative} for a comprehensive review of multilateral bargaining and \citet{ray2015coalition} for a review on coalition formation.} In this paper, we extend the \citet{baron1989bargaining} model under closed rules and add the mayor's first move and political partisanship. Partisanship here is different from \citet{choate2019partisan} since we consider partisanship as a zero-sum effect that is positive for politicians aligned with the mayor and negative otherwise. This paper investigates the often overlooked effect of legislature size on the bargaining equilibrium strategies.

Second, this paper contributes to the literature on the effects of larger legislatures in service provision. Following the contribution by \citet{Weingastetal1981}, the "law of $1/n$" posits that larger legislatures increase the legislative pressure for providing targeted public goods to local supporters at the expense of the entire polity \citep{freire2021legislature}. The corollary of the "law of $1/n$" is that increasing legislature size promotes excessive public spending.\footnote{\citet{PrimoSnyder2008} show that legislature size may result in positive or negative expenditures, depending on factors such as the degree that local public goods can exclude non-locals or deadweight costs associated with taxation.} In the US metaphor, this would explain significant infrastructure spending in smaller districts. However, the "law of $1/n$" literature rarely mentions the relationship between public goods provision and welfare. This paper is one of the first to study how the number of local legislators affects welfare. Our model provides a theoretical expectation based on the chances of electing a politician aligned with the mayor. The empirical estimates confirm our theoretical predictions, with the additional fact that legislature size has a null effect on expenditures in Brazil.

Third, our article also contributes to a growing literature on municipal partisan politics. Most papers in this area estimate the effects of the mayor's party on political outcomes. For instance, \citet{gerber2011mayors} and \citet{de2016mayoral} find differences in governance when comparing Democratic and Republican mayors. Specifically about Brazil, \citet{gouvea2021partisanship} shows that social policy expenditure increase with left-wing mayors. \citet{frey2020larger}, using discontinuities in city council size after the 2009 constitutional amendment that reverted the thresholds we study here, shows that changes in council size lower the support for candidates backed by the municipal mayors. Our paper reinforces the view that partisanship matters locally, even in countries with weak ideological linkage.

Finally, this paper contributes to a growing literature on the causes and effects of legislature sizes. Most papers on legislature size study the trade-off between representation and efficiency \citet{Stigler1976, muzzio1989size, frederick2008people, rush2013conclusion, jacobs2015explaining}. However, a few new papers explore the effects of legislature size on lawmaking and partisanship \citep{rush2013conclusion, kirkland2014chamber}. For instance, \citet{kirkland2014chamber} shows that larger legislatures make it harder for politicians to know each other and collaborate, making them rely more heavily on party cues and committee structures. In considering the causes of changes in legislature sizes, \citet{jacobs2015explaining} show that population shifts and economic crises cause enlargement and shrinkage, respectively. \citet{marland2019fewer} questions the economic crises mechanism, showing that political elites shrink legislatures mostly to signal to the public the priority of their austerity agenda. In this paper, we show that legislature size affects welfare by changing the partisan composition of legislatures.

@@

The design of institutions is crucial for long term welfare \citep{north1989constitutions, buenodemesquita2005logic, acemoglu2012nations} and democratic consolidation \citep{lijphart2018institutional, gandhi2019institutional}. One particular problem that each polity faces is determination of the size of its legislative bodies. The literature is extensive, and modern attempts to provide a solution to the legislature size problem started as early as in the eighteenth century, with the Federalists versus Anti-Federalists debates \citep{Stigler1976, muzzio1989size}.

The legislature size literature highlights that the number of legislators causes tension between representation and efficiency \citep{taagepera1972size, Stigler1976, muzzio1989size, frederick2008people, rush2013conclusion, jacobs2015explaining}. Larger legislatures improve representation as the election thresholds decrease and the ratio of population-per-politician decreases. This facilitates the electoral success of ethnic and political minorities \citep{allenrepresentation}. Conversely, the improvements in representation are countered by the decline in efficiency in larger legislatures. The enlarged size makes it harder for members to coordinate and collaborate without partisanship and legislative structures \citep{rush2013conclusion, kirkland2014chamber}. It also increases the potential number of veto players, decreasing the policymaking efficiency \citep{tsebelis2002veto}. Balancing these two incentives is crucial for the optimal legislative composition \citep{taagepera1972size, auriol2012optimal}.

However, the legislature size scholarship has overlooked two questions: first, does legislature size matter in contexts where the legislative bodies are not the pivotal decision-maker? And second, what is the ultimate effect of legislature size on welfare? On the one hand, legislature size's effects on welfare represent a significant gap in the political economy literature. It is likely generated by the predominance of the "law of 1/n" scholarship, which emphasized larger legislatures' overspending consequence. Since the seminal paper by \citet{Weingastetal1981}, 30 papers have systematically tested the effects of legislature size on government spending \citep{freire2021legislature}. Of these papers, only \citet{lewis2019legislature} investigated its effects on welfare, showing that larger legislatures have a negative welfare effect.\footnote{\citet{koppel2004public} also shows, using a repeated game where the proposal prerogative rotates in a legislature, that larger legislatures increase the chances of non-cooperative outcomes, lowering public goods provision.} Moreover, \citet{frey2020larger} finds, for the Brazilian case after the 2009 legislature size and salaries constitutional amendment, a null effect in service provision from 2012 to 2016. These findings suggest that the legislature size and welfare nexus need a dedicated theoretical and empirical assessment.

On the other hand, albeit the division of power, in most mayor-council (or presidential) systems, the Executive dominates the service provision. For instance, while the New York Times cited the New York City Mayor Bill De Blasio 2,084 times in 2021, it mentioned the 49 seating city councilors 440 times in the same period. To avoid the COVID crisis when the Executive was in the spotlight, we expanded the search for the past five years. In this new search, Bill De Blasio was mentioned 8,978 while the 49 seating city councilors were mentioned 2,046 times.\footnote{These figures were computed using the Dow Jones Factiva Database in November 2021.} The mayor was cited, on average, over four times more than all the city councilors combined. This Executive dominance has implications for policymaking, as coalitions and affinity with the Executive are going to be crucial for service provision \citep{buenodemesquita2005logic}.

In a mayor-council setting, consider a strategic interaction between a mayor ($M$) and $N \geq 3$ city councilors indexed in $i \in \{1, 2, \cdots, N \}$. The mayor has the prerogative of proposing a vector of policies on which city councilors vote. The provision of public goods is then a bargain between the mayor and the city council. Policy proposals are a combination of public goods provision $g$, rents for the mayor $r$, and a vector of transfer for city councilors \textbf{x}.

@@

We now consider a situation that combines both partisan and non-partisan legislature concerns. Each councilor has a party affiliation. Party affiliations are mutually exclusive, and if a city councilor is aligned with the mayor's party, then she belongs to the government coalition $G \subset \{1, 2, \cdots, N \}$. Otherwise, the city councilor belongs to the opposition $O = \{1, 2, \cdots, N \} \setminus G$. The mayor's policy generates a political value of $p > 0$ for the councilors, and whether it adds or subtracts from the councilor's utility depends on the party affiliation of the councilor. If councilors reject the mayor's proposal, no public goods are provided ($p=0$). Moreover, in equilibrium, we assume that city councilors would prefer to receive transfers instead of providing public goods.\footnote{In the Full Supplementary Materials (available in the paper's GitHub page), we consider a version of the model where public goods provision depends on the party affiliation of the proposer so that it affects choices in any potential history of the game. The equilibrium public goods provision level is even higher than in the model presented here, as city councilors prefer to approve the mayor's proposal than run the risk of an opposition member becoming the proposer.}

@@

Second, more seats could increase the competitiveness in the election, making politicians provide more services \citep{besley2010political}. In a political system with around thirty parties and a sample average of almost ten parties in the municipalities we study, candidates per seat can increase exponentially. According to the Brazilian electoral law, each party can sponsor 1.5 times the number of seats as candidates. If a party forms a pre-electoral coalition, this number increases to two times the number of seats \citep{coneglian2008lei}. An election in a small town, with ten councilors, may end up with 600 candidates: (30 parties) x (two times the number of seats) x (ten seats). Therefore, an alternative hypothesis to explain the increased service provision is that larger council size increase the competitiveness of the election.\footnote{Despite the common assumption that competitiveness improve welfare, \citet{boulding2014political} and \citet{gottlieb2019countervailing} can increase budgetary pressure, lower social policy, lead to bargaining inefficiencies.}

@@
